preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2020-09-29 Manuscript/manuscript.Rmd
uncommented figures for bioarxiv submission

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state (BroadHMM), histone modification (HM), and transcription factor binding site (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary region prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites improve prediction performance, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD*-predicted boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://dozmorovlab.github.io/preciseTAD/ and Bioconductor.

@@

<!--
![**Resolution-specific data construction and feature engineering for random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals using *shifted binning* (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD (or loop) boundary. (B) Four types of associations between bins (blue dashed lines) and genomic annotations (green shapes) were considered to build the predictor space including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and $log_{2}$ distance (Distance).](Figures/figure1.png)
-->

@@

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb      |                         8,052                       |                             15,468                           |                535,363               |           0.03         |
|     Arrowhead         |       10   kb     |                         7,676                       |                             14,253                           |                267,682               |           0.05         |
|     Arrowhead         |       25   kb     |                         4,670                       |                             8,363                            |                107,073               |           0.08         |
|     Arrowhead         |       50   kb     |                         2,349                       |                             4,224                            |                 53,537               |           0.08         |
|     Arrowhead         |      100   kb     |                         1,031                       |                             1,883                            |                 26,768               |           0.07         |
|        Peakachu       |       10   kb     |                        16,185                       |                          21,421                              |                267,682               |        0.14            |
Table: Domain boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on GM12878.

@@

<!--
![**A machine learning framework for building domain boundary region prediction models.** Step 1 employs a range of feature engineering techniques to define the predictor matrix $A_{N \times (p)}$, where $N$ is the number of genomic bins, $p$ is the number of genomic annotations, $i$ is a holdout chromosome. The response vector $Y_N$ is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$). Step 2 reserves the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 applies a resampling technique to the training data to address the class imbalance. Step 4 trains the random forest model and performs 3-fold cross-validation to tune the mtry parameter. Finally, step 5 validates the model on the separate test data composed of the binned data from the holdout chromosome $i$ and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png) 
-->

@@

<!--
![**Determining optimal data level characteristics for building TAD boundary region prediction models on GM12878.** Averaged balanced accuracies are compared across resolution, within each predictor-type: Signal, OC, OP, and Distance, and across resampling techniques: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate standard deviation from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)
-->

@@

<!--
![**SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predict TAD and loop boundaries in GM12878.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using histone modifications (HM), chromatin states (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the mean cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and the corresponding average area under the curves (AUCs) when training and testing on GM12878 data (blue, Arrowhead ground truth; red, Peakachu ground truth) versus training on K562 and testing on GM12878 data (black, dashed). The curves represent the average sensitivities and specificities across each holdout chromosome. The shaded areas around each curve represent 1 standard deviation from the average.](Figures/figure4.png)
-->

@@

<!--
![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR, highlighted yellow), a cluster of bases with high probability of being a boundary, and the corresponding signal profiles of CTCF, RAD21, SMC3, and ZNF143. (C, D) Signal profile plots comparing the strength of CTCF, RAD21, SMC3, and ZNF143 binding around Arrowhead-called boundaries (blue, C), Peakachu loop boundaries (red, D) vs. *preciseTAD*-predicted boundaries (green).](Figures/figure5.png) 
-->

@@

<!--
![**preciseTAD-predicted boundaries are closer to CTCF sites and more conserved across cell lines**. (A) Log2 genomic distance distribution from called and predicted boundaries to nearest CTCF sites. The p-values are from the Wilcoxon Rank Sum test. (B-E) Venn diagrams illustrating the levels of conservation (overlap) between domain boundaries for GM12878 (red) and K562 (blue) cell lines identified by Arrowhead (B), Peakachu (D), *preciseTAD*-predicted boundaries using Arrowhead- and Peakachu-trained models () Arrowhead-trained *preciseTAD* models (C), Peakachu boundaries themselves (C, E, respectively). Boundaries involving Arrowhead/Peakachu were flanked by 5 kb/10 kb, respectively.](Figures/figure6.png) 
-->

@@

# Figure legends

@@

**Figure 6. preciseTAD-predicted boundaries are closer to CTCF sites and more conserved across cell lines**. (A) $log_{2}$ genomic distance distribution from called and predicted boundaries to nearest CTCF sites. The p-values are from the Wilcoxon Rank Sum test. (B-E) Venn diagrams illustrating the levels of conservation (overlap) between domain boundaries for GM12878 (red) and K562 (blue) cell lines identified by Arrowhead (B), Peakachu (D), and *preciseTAD*-predicted boundaries using (C) Arrowhead- and (E) Peakachu-trained models. Boundaries involving Arrowhead/Peakachu were flanked by 5 kb/10 kb, respectively.

>>>>

We present *preciseTAD*, an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on high-resolution chromatin state (BroadHMM), histone modification (HM), and transcription factor binding site (TFBS) data to predict low-resolution boundaries. We introduce a systematic pipeline for building the optimal boundary region prediction classifier. We found that spatial associations (linear distance) between boundaries and annotations perform best, transcription factor binding sites improve prediction performance, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and partitioning around medoids (PAM) to detect genome annotation-guided boundary regions and points at a base-level resolution. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD*-predicted boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on https://bioconductor.org/packages/preciseTAD.

@@

![**Resolution-specific data construction and feature engineering for random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals using *shifted binning* (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD (or loop) boundary. (B) Four types of associations between bins (blue dashed lines) and genomic annotations (green shapes) were considered to build the predictor space including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and $log_{2}$ distance (Distance).](Figures/figure1.png)

@@

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb      |                         8052                       |                             15468                           |                535363               |           0.03         |
|     Arrowhead         |       10   kb     |                         7676                       |                             14253                           |                267682               |           0.05         |
|     Arrowhead         |       25   kb     |                         4670                       |                             8363                            |                107073               |           0.08         |
|     Arrowhead         |       50   kb     |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     Arrowhead         |      100   kb     |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        Peakachu       |       10   kb     |                        16185                       |                          21421                              |                267682               |        0.14            |
Table: Domain boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on GM12878.

@@

![**A machine learning framework for building domain boundary region prediction models.** Step 1 employs a range of feature engineering techniques to define the predictor matrix $A_{N \times (p)}$, where $N$ is the number of genomic bins, $p$ is the number of genomic annotations, $i$ is a holdout chromosome. The response vector $Y_N$ is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$). Step 2 reserves the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 applies a resampling technique to the training data to address the class imbalance. Step 4 trains the random forest model and performs 3-fold cross-validation to tune the mtry parameter. Finally, step 5 validates the model on the separate test data composed of the binned data from the holdout chromosome $i$ and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png)

@@

![**Determining optimal data level characteristics for building TAD boundary region prediction models on GM12878.** Averaged balanced accuracies are compared across resolution, within each predictor-type: Signal, OC, OP, and Distance, and across resampling techniques: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate standard deviation from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)

@@

![**SMC3, RAD21, CTCF, and ZNF143 transcription factors accurately predict TAD and loop boundaries in GM12878.** (A) Barplots comparing performances of TAD (Arrowhead) and loop (Peakachu) boundary prediction models using histone modifications (HM), chromatin states (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent 1 standard deviation from the mean cross-validated accuracy across each holdout chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 most predictive chromosome-specific TFBSs. The columns represent the holdout chromosome excluded from the training data. (D) Receiver operating characteristic (ROC) curves and the corresponding average area under the curves (AUCs) when training and testing on GM12878 data (blue, Arrowhead ground truth; red, Peakachu ground truth) versus training on K562 and testing on GM12878 data (black, dashed). The curves represent the average sensitivities and specificities across each holdout chromosome. The shaded areas around each curve represent 1 standard deviation from the average.](Figures/figure4.png)

@@

![**preciseTAD boundaries are enriched for known molecular drivers of 3D chromatin.** (A) The location of Arrowhead-called TAD boundaries (blue) vs. *preciseTAD*-predicted TAD boundaries (green) on GM12878 data (chr14:50085000-50800000). The black line represents the predicted probability of each base being a TAD boundary. (B) A zoomed-in portion of the genome shows the preciseTAD boundary region (PTBR, highlighted yellow), a cluster of bases with high probability of being a boundary, and the corresponding signal profiles of CTCF, RAD21, SMC3, and ZNF143. (C, D) Signal profile plots comparing the strength of CTCF, RAD21, SMC3, and ZNF143 binding around Arrowhead-called boundaries (blue, C), Peakachu loop boundaries (red, D) vs. *preciseTAD*-predicted boundaries (green).](Figures/figure5.png)

@@

![**preciseTAD-predicted boundaries are closer to CTCF sites and more conserved across cell lines**. (A) Log2 genomic distance distribution from called and predicted boundaries to nearest CTCF sites. The p-values are from the Wilcoxon Rank Sum test. (B-E) Venn diagrams illustrating the levels of conservation (overlap) between domain boundaries for GM12878 (red) and K562 (blue) cell lines identified by Arrowhead (B), Peakachu (D), *preciseTAD*-predicted boundaries using Arrowhead- and Peakachu-trained models () Arrowhead-trained *preciseTAD* models (C), Peakachu boundaries themselves (C, E, respectively). Boundaries involving Arrowhead/Peakachu were flanked by 5 kb/10 kb, respectively.](Figures/figure6.png)

@@

<!--
# Figure legends

@@

**Figure 6. preciseTAD-predicted boundaries are closer to CTCF sites and more conserved across cell lines**. (A) $log_{2}$ genomic distance distribution from called and predicted boundaries to nearest CTCF sites. The p-values are from the Wilcoxon Rank Sum test. (B-E) Venn diagrams illustrating the levels of conservation (overlap) between domain boundaries for GM12878 (red) and K562 (blue) cell lines identified by Arrowhead (B), Peakachu (D), and *preciseTAD*-predicted boundaries using (C) Arrowhead- and (E) Peakachu-trained models. Boundaries involving Arrowhead/Peakachu were flanked by 5 kb/10 kb, respectively.
-->
